---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo = TRUE}
data1 <- read.csv(file = "activity.csv", header = T, sep = ",")
```
## What is mean total number of steps taken per day?
```{r echo = TRUE, fig.path = 'figure/'}
stepsDay <- by(data1$steps, data1$date, mean, na.rm = T)
hist(stepsDay, main = "Steps per day", xlab = "Mean steps per day")
stepMean <- mean(stepsDay, na.rm = T)
stepMedian <- median(stepsDay, na.rm = T)
```
Mean total number of steps is `r round(stepMean,1)`.

Median total number of steps is `r round(stepMedian,1)`.

## What is the average daily activity pattern?
```{r echo = TRUE, fig.path = 'figure/'}
avgSteps <- by(data1$steps, data1$interval, mean, na.rm = T)
plot(unique(data1$interval), avgSteps, type = "l",
     ylab = "Average steps per interval", 
     xlab = "5-minute interval",
     main = "Time series plot")
maxStep <- data1$interval[which.max(avgSteps)]
```
The interval `r maxStep` contains the maximum average number of steps.

## Imputing missing values
```{r echo = TRUE}
totalNa <- sum(is.na(data1$steps) | is.na(data1$date) | is.na(data1$interval))
stepNa <- sum(is.na(data1$steps))
dateNa <- sum(is.na(data1$date))
intervalNa <- sum(is.na(data1$interval))
```
The total number of missing values is `r totalNa`; there are `r stepNa`, `r dateNa`
and `r intervalNa` missing values in steps, date and interval columns, accordingly.

I will be using the median step number for the time interval to fill in the missing 
values.

```{r echo = T, fig.path = 'figure/'}
data2 <- data1
naVec <- is.na(data2$steps)

medSteps <- by(data2$steps, data2$interval, median, na.rm = T)
names(medSteps) <- unique(data2$interval)

data2$steps[naVec] <- medSteps[as.character(data2$interval[naVec])]

stepsDay2 <- by(data2$steps, data2$date, mean)

hist(stepsDay2, main = "Steps per day after imputation", xlab = "Mean steps per day")
stepMean2 <- mean(stepsDay2)
stepMedian2 <- median(stepsDay2)
```
Mean total number of steps after imputation is `r round(stepMean2,1)`.

Median total number of steps after imputation is `r round(stepMedian2,1)`.

These values differ from the estimate without imputation, mean is lower by 
`r stepMean - stepMean2` steps, and median is lower by `r stepMedian - stepMedian2` steps.

## Are there differences in activity patterns between weekdays and weekends?
```{r echo = TRUE, fig.path = 'figure/'}
Sys.setlocale("LC_TIME", "en_US") # for compatibility
data2$dayType <- factor(ifelse(
    weekdays(
        as.Date(data2$date, "%Y-%m-%d"), abbreviate = T
    ) %in% c("Sat", "Sun"), "weekend", "weekday"
))

library(ggplot2)
library(data.table)
data2 <- as.data.table(data2)

ggplot(data2[,list(steps = mean(steps)), by=list(interval, dayType)]) +
    aes(interval, steps)+
    geom_line()+
    facet_grid(dayType ~ .)+
    xlab("Interval")+
    ylab("Mean steps per day")

```

So, the activity patterns are different.